\newacronym{foss}{FOSS}{Free Open Source Software}
\newacronym{ict}{ICT}{Information and Communications Technology}
\newacronym{eden}{EDEN}{Emergency Development ENvironment}
\newacronym{ict4d}{ICT4D}{Information and Communication Technologies for Development}


\newglossaryentry{ICT}
{
    name=Information and communications technology,
    description={Information and communications technology: A term for the mix of physical devices and software that mix analog and digital techniques to allow communication, recall and storage of data.}
}
 
\newglossaryentry{FOSS}
{
    name=Free Open Source Software,
    description={Free Open Source Software: A method of developing software that does not charge for use of the software. The software's source code is also freely distributed, allowing users to modify their copy of the software or suggest changes to the organization developing it.}
}

\newglossaryentry{python}
{
    name=python,
    description={Python is an interpreted, high-level, general-purpose programming language.}
}

\newglossaryentry{OpenStreetMap}
{
    name=OpenStreetMap,
    description={An open community of volunteers that maintain data about roads, trails, cafés, railway stations, and much more, all over the world.}
}
\newglossaryentry{web framework}
{
    name=web framework,
    description={A heterogeneous set of software that allow a programmer to efficiently write and manage providing a service over the Internet. This could be a website or a mobile app (often it is both) or a go-between for other services. Examples include web2py and Django in Python or Phoenix in Elixir.}
}

\newglossaryentry{PostgreSQL}
{
   name=PostgreSQL,
   description={PostgreSQL is a powerful, open source object-relational database system with over 30 years of active development that has earned it a strong reputation for reliability, feature robustness, and performance.}
   % https://www.postgresql.org/
}

% https://en.wikipedia.org/wiki/MySQL
\newglossaryentry{MySQL}
{
   name=MySQL,
   description={MySQL is an open-source relational database management system (RDBMS). Its name is a combination of "My", the name of co-founder Michael Widenius's daughter, and "SQL", the abbreviation for Structured Query Language. MySQL is a component of the LAMP web application software stack (and others), which is an acronym for Linux, Apache, MySQL, Perl/PHP/Python. MySQL is used by many database-driven web applications, including Drupal, Joomla, phpBB, and WordPress.}
}

% http://eden.sahanafoundation.org/wiki/WikiStart#WhatisSahanaEden
\newglossaryentry{EDEN}
{
   name=EDEN,
   description={The Emergency Development ENvironment, developed by the Sahana Foundation, is an Open Source Humanitarian Platform which can be used to provide solutions for Disaster Management, Development, and Environmental Management sectors.}
}

\newglossaryentry{ICT4D}
{
   name=Information and Communication Technologies for Development,
   description={The movement to use digital technologies to serve the needs of the underprivileged and those living in the developing world. Criticized for being directed more at the developed worlds' love of technology than the actual needs of those it purports to serve. Also called digital humanitarianism.}
}

% \newglossaryentry{}
% {
%    name=,
%    description={}
% }